---
title: "36-402 Homework 5"
author: "Platt Yoon"
subtitle: Due Tuesday, Feburary 26th by 5:59pm
output:
  pdf_document: default
  html_document:
    df_print: paged
urlcolor: blue
---

# Question 1

$E[T] = \sum_t tP(T=t) = 0*P(T=0) + 1*P(T =1) = P(T=1)$

**Since we are only using binary varibles, we know that $T^2= T$ since the square of both 0 and 1 are both 0 and 1 respectively.**

$Var[T] = E[T^2] -(E[T])^2 = P-P^2$ **(based on E[T] computed above and the fact that $T^2=T$)** 
$= p(1-p)$

# Question 2

## part a

$E[S_q] = E[\sum_{i=1}^qT_i] = \sum_{i=1}^qE[T_i] = \sum_{i=1}^qp = q*p$

$var[s_q] = var[\sum_{i=1}^qT_i] = \sum_{i=1}^qvar[T_i] = \sum_{i=1}^qp(1-p) = qp(1-p)$

## part b

**$P(S_q=s)$ can be split into two parts in which when s of the $T_i =1$ we have $p^s$ and when $T_i = 0$ we have $(1-p)^{q-s}$. With this in mind there are a combination of $\binom{q}{s} T_i$**

$P(S_q = s) = \binom{q}{s}p^s(1-p)^{q-s} = \frac{q!}{s!(q-s)!}p^s(1-p)^{q-s}$


# Question 3

## part a

**Since from above we have that $E[S_q] = qp$ and that $p = \mu/q$**

$E[S_q] = qp = q(\mu/q) = \mu$

## part b

**Since from above we have that $var[S_q] = qp(1-p)$ and that $p = \mu/q$**

$var[S_q] = qp(1-p) = q\frac{\mu}{q}(1-\frac{\mu}{q}) = \mu(1-\frac{\mu}{q})$

**Since the $\frac{\mu}{q}$ term goes to 0 as q goes to infinity, we are left with **

$\mu*1 = \mu$

## part c

**From the previous problem we know that$P(S_q = s) = \binom{q}{s}p^s(1-p)^{q-s}$**

**From the question we know that $p = \mu/q$**

$P(S_q = s) = \binom{q}{s}p^s(1-p)^{q-s} = \binom{q}{s} (\frac{\mu}{q})^s(1-\frac{\mu}{q})^{q-s}$

**From here we will split the equation into two parts**

**The first part we will only consider $\binom{q}{s} (\frac{\mu}{q})^s$**

$\binom{q}{s} (\frac{\mu}{q})^s  = \frac{q(q-1)....(q-s+1)}{s!}\frac{\mu^s}{q^s} = \frac{\mu}{s!}\prod_{k=0}^{s-1} \left(1 - \frac{i}{q} \right)$

**As q goes to infinity the term converges to 1 (since $\frac{k}{q}$goes to 0 as q goes to infinity, we have **

$\frac{\mu^s}{s!}$

For the second part we have 

$(1-\frac{\mu}{q})^{q-s} = (1-\frac{\mu}{q})^q(1-\frac{\mu}{q})^{-s}$

**From the limiting factor we know that as q goes to infinity $(1+\frac{x}{q})^q = e^x$ where $x=-\mu$ Therefore,**

$(1-\frac{\mu}{q})^q= e^{-\mu}$ 

**Moreover, $(1-\frac{\mu}{q})^{-s}$ converges to 1 since s is fixed and as mentioned above as q goes to infinity, the term $\frac{\mu}{q}$ goes to 0. Multiplying the limits, we have**

$P(S_q = s) -> \frac{\mu^s e^{-\mu}}{s!}$


## part d

**From question 3a and 3b, we know that this is a poisson distribution and we have already computed the mean/variance of $S_q$ as well as that $p = \mu/q$**

**From Q3(a) we have $E[S_q] = pq = q(\mu/q) = \mu$**

**From Q3(b) we have $var[S_q] = qp(1-p) = q\frac{\mu}{q}(1-\frac{\mu}{q}) = \mu(1-\frac{\mu}{q})$ and since as q goes to infinity this converges into $\mu*1 = \mu$**

**Since $S_q$ converges in distribution to a poisson($\mu$) random variable and the limiting distribution is the Poisson pmf, the limiting mean and variance are $\mu$ and thus $Y~Poisson(\mu)$**

## part e

**We have many trials/oppurtunities(q) while each trial has a small probability p of an event(thus each event is rare) and we count the total number of events $S_q = \sum_{i=1}^q T_i$. Therefore, poisson counts model the total number of occurrences when each individual oppurtunity has little chance to produce an event but there are many opportunities overall.**

# Question 4

## part a

$L(\mu_1....\mu_n) = \sum_{i=1}^nlogP(Y_i=y;\mu_i)$

**Since we know that Yi~Pois($\mu_i$) and each the $\mu_i$ terms are iid, and from the previous problem Q3(d) we know that **

$P(Y_i=y_i;\mu_i) = \frac{\mu^y_i e^{-\mu_i}}{y_i!}$ (after substitution from s to $y_i$)

**By the log properties we get**

$L= log(\mu_i^{y_i}) + log(e^{-\mu_i})-log(y_i!) = y_ilog(\mu_i) -\mu_i - log(y_i!)$

**Therefore when we plug this back into the log-liklihood function we get:**

$L(\mu_1....\mu_n) = (\sum_{i=1}^ny_ilog(\mu_i))-(\sum_{i=1}^n \mu_i)-(\sum_{i=1}^nlog(y_i!))$

**Friom this we can see that the last term is a function of $y_i$ and does not depend on $\mu_i$ while the second term is a function of $\mu_i$ but not of $y_i$ and the first term is proportional to $y_i log(\mu_i)$**

## part b

**When we do derivative in terms of $\mu_i$ to the log-likelihood function and we set it to 0,**

\[
\frac{\partial \ell}{\partial \mu_i} = 0
\]

**After wards we solve for $\hat\mu_i$ and since we know that $P(Y_i=y_i;\mu_i) = \frac{\mu^y_i e^{-\mu_i}}{y_i!}$ when we plug it back into the loss function and solve it from there, we get  **

\[
\frac{\partial \ell}{\partial \mu_i}
= \frac{\partial}{\partial \mu_i}
\left[
\log \left(
\frac{e^{-\mu_i} \mu_i^{y_i}}{y_i!}
\right)
\right]
\]

\[
= \frac{\partial}{\partial \mu_i}
\left[
y_i \log \mu_i - \mu_i - \log(y_i!)
\right]
\]

\[
= \frac{y_i}{\mu_i} - 1
\]

\[
\frac{y_i}{\mu_i} - 1 = 0
\]

\[
\Rightarrow \frac{y_i}{\mu_i} = 1
\]

\[
\Rightarrow \hat{\mu}_i = y_i
\]

# Question 5

## part a

**If we let N be the number of people in the city(very large) on a given day, we define indicator variables as: **

\[
T_i =
\begin{cases}
1, & \text{if person } i \text{ dies on a given day} \\
0, & \text{otherwise}
\end{cases}
\qquad (i = 1, \dots, N)
\]

**The total deaths that day can be shown as: **

\[
S = \sum_{i=1}^{N} T_i.
\]

**Here, each individual death is rare, and there are many opportunities for the event since N is huge and the total count is a sum of rare events which exactly matches what the poisson model is.**

## part b

**In Poisson regression, the mean parameter must be positive: $(\mu > 0)$. If you model $(\mu)$ directly as a linear function of predictors, the result could become negative, which is impossible for a Poisson mean.**

**Using a log link resolves this issue by changes the range such that:**
\[
\mu \in (0, \infty)
\quad \longrightarrow \quad
\log \mu \in (-\infty, \infty).
\]

**Now we know that the predictor can take any real value and transforming it back to its original form we have: **
\[
\mu = \exp(\eta),
\]
**which guarantees $(\mu > 0)$.**


**Moreover, in the log-likelihood, it is natural whereas shown down below for $Y_i ~ Pois(\mu_i)$:**

\[
\ell(\mu_i) = \log P(Y_i = y_i)
= y_i \log(\mu_i) - \mu_i - \log(y_i!).
\]

**Thus, the log-likelihood depends on $(\mu_i)$ through $(\log(\mu_i))$, making $(log \mu_i)$ the mathematically convenient scale to model linearly.**

## part c

**We know that $log(\mu_i)$ is an additive function of some predictor variables and if we put it terms of the additive model we get:**

$log(\mu) = \alpha + \sum_{j=1}^d f_j(x_j)$ **for predictor variables $x_1...x_d$**

**From this if we exponent each side we have:**

$\mu = exp(\alpha+ \sum_{j=1}^d f_j(x_j)) = e^\alpha \prod_{j=1}^de^{f_j(x_j)}) = y_0\prod_{j=1}^de^{g_j(x_j)}$

**Therefore the additive in $log\mu$ become multiplicative in $\mu$**



# Question 6

## part a

```{r,echo=FALSE}
library(gamair)
data(chicago)
fit <- smooth.spline(chicago$time, log(chicago$death)) 
plot(chicago$time, chicago$death,
     pch = 16, cex = 0.5,
     xlab = "time (days since 1993-12-31)",
     ylab = "death",
     main = "death as a function of time")

lines(chicago$time,
      exp(predict(fit)$y),
      col = "red",)

legend("topright",
       legend = "Spline fit",
       col = "red",
       pch = c(16, NA),
       bty = "n")
```


## part b
```{r,echo=FALSE}
library(knitr)
chicago$date <- as.Date("1993-12-31") + chicago$time
imax <- which.max(chicago$death)
idx4 <- (imax - 1):(imax + 2)
idx4 <- idx4[idx4 >= 1 & idx4 <= nrow(chicago)]
out <- chicago[idx4, c("date", "death")]
kable(out, caption = "Four consecutive high-death days")
```
**The 4 dates are 1995-07-13,1995-07-14,1995-07-15,and 1995-07-16**

## part c

**we fit the spline to log(death) since deaths often have underlying seasonal and long-term patterns and are often multiplicative and thus taking the logarithm converts the multiplicative effects into additive ones and helps stablize the varince. Moreover, the log transformation guarantees that after exponenting the values, the fitted values remain positive which is necessary for death counts since they are always going to be positive. We still present the plot on the original death scale since death counts are directly interpretable.**

# Question 7 

## part a

```{r,echo=FALSE,message=FALSE}
library(mgcv)

m7 <- gam(death ~ s(pm10median) + s(o3median) + s(so2median) + s(tmpd),
          family = "poisson",
          data = chicago)

par(mfrow = c(2,2), mar = c(4,4,2,1))

plot(m7,
     residuals = TRUE,
     pch = 16,
     cex = 0.35, 
     main = "With Residuals")

par(mfrow = c(2,2), mar = c(4,4,2,1))

plot(m7,
     residuals = FALSE,
     lwd = 2,
     main = "Without Residuals")
```
**The vertical axis on each partial response plot is $f_j(x)$ which is measured in log mean death scale. This means that a vertical value of 0 has no effect while a vertical value of 0.1 means that it multiplies expected death by $e^{0.1} = 1.105$ and a vertical value of -0.1 multiplies the expected death by $e^{-0.1} = 0.905$. Therefore the units are the additive changes in log expected deaths or the multiplicative effects on expected death counts**

**In terms of the partial response functions, pm10 is fairly flat at low values and gradually increases with higher values with a slight upward curvature at the right tale. This means that at low-t- moderate pm10 levels the effect is small and at higher pm10 levels the smooth increases steadily meaning higher particulate pollution is associated with higher expected deaths. For ozone, the graph shows a fairly flat shape in the middle with downward curvature at higher levels with larger confidence bands in extremes. This shows that ozone overall has little association with mortality in lower quantities and for higher quantities, since the confidence bands widen substantially, suggests the decline at very high ozone values are uncertain and likely due to sparse data. The so2 shows a nearly flat shape throughout with only slight curvature at high values with wide uncertainty at higher values. This shows little evidence that sulfur dioxide concentrations are associated with death and any effect appears small and uncertain. Lastly, temperature show a clear non-relationship where there is a slight elevation for low temperatures and a decrease at moderate termperatures and a sharp increase at very high temperatures. This shows that mortality is lowest at moderate termperatures and increases at extreme heat or extreme cold.**

## part b

```{r,echo=FALSE}
#From lines.mdl.NApadded in syllabus
lines.mdl.NApadded <- function(mdl, data=chicago, subset=1:nrow(data),
                               xcol="date",
                               ...) {
    # Invoke predict() for mdl on data, using the fact that the default behavior
    # and tell predict to return NA for any
    # row of data where variables are missing, rather than omitting that row
      # Not all predict() methods recognizd the na.action argument, but
      # predict.lm(), predict.glm() and predict.gam() all do so
      # Note that fitted() does not do this!
    preds.padded.with.NAs <- predict(mdl,
                                     newdata=data[subset,],
                                     type="response",
                                     na.action=na.pass)
    lines(x=data[subset, xcol],
          y=preds.padded.with.NAs,
          ...)
}

plot(chicago$date, chicago$death,
     pch = 16, cex = 0.5,
     col = "grey40",
     xlab = "date", ylab = "death",
     main = "fitted values as a function of time")

lines.mdl.NApadded(m7, data = chicago, xcol = "date",
                   col = "red", lwd = 2)
legend("topright",
       legend = c("Observed deaths", "GAM fitted mean"),
       col = c("grey40", "red"),
       pch = c(16, NA),
       lwd = c(NA, 2),
       bty = "n")
```


## part c

**Yes the large outliers are still present. Even after the Poisson GAM, the observed death counts continue to show several extreme spikes around 1997. The Gam fitted mean is smooth so it captures gradual seasonal and long-term variation, however, it does not capture extreme outliers and thus the extreme days remaining visible in the data and aren't significantly better and seem to be the same.**

# Question 8

## part a

**In this function, the argument window determines how many lagged observations are included when computing the moving average. The loop runs from t=0 to t=window which produces window+1 shifted version of the series that are summed up and then divided by window+1. Therefore, k is the number of days being averaged and the relationship is k= window +1. Moreover, the function returns the average value of a variable over today and the previous k days by summing multiple lagged versions of the series and dividing by the number of terms. From the loop, we get window +1 slices of x and each slice aligns values so that when added, they combine the current observation with its previous lags and after these aligned y values are accumulated, the fuction divides window+1 which produces their mean. **

## part b

```{r,echo=FALSE,message=FALSE}
# lag.mean from the handout
lag.mean <- function(x, window) {
  n <- length(x)
  y <- rep(0, n - window)
  for (t in 0:window) {
    y <- y + x[(t+1):(n - window + t)]
  }
  return(y / (window + 1))
}

k <- 3  
new.chicago <- chicago[(k+1):nrow(chicago), ]
avg.cols <- c("tmpd", "pm10median", "o3median", "so2median")
for (nm in avg.cols) {
  new.chicago[[nm]] <- lag.mean(chicago[[nm]], k)
}

expected_rows <- nrow(chicago) - 3
set.seed(1)
pairs <- data.frame(
  row = sample(1:nrow(new.chicago), 3),
  col = sample(avg.cols, 3, replace = TRUE),
  stringsAsFactors = FALSE
)
tol <- 1e-10
check <- data.frame(row = integer(), col = character(),
                    computed = numeric(), manual = numeric(),
                    diff = numeric(), pass = logical(),
                    stringsAsFactors = FALSE)

for (j in 1:nrow(pairs)) {
  i <- pairs$row[j]
  nm <- pairs$col[j]
  idx.today <- i + k  
  manual <- mean(chicago[[nm]][(idx.today - k):idx.today], na.rm = TRUE)
  computed <- new.chicago[[nm]][i]
  diff <- computed - manual
  check <- rbind(check, data.frame(
    row = i,
    col = nm,
    computed = computed,
    manual = manual,
    diff = diff,
    pass = abs(diff) < tol
  ))
}

kable(check, digits = 6, caption = "Spot-check of lag.mean averaging")
```

**the new data should have 3 rows shorter than the original which is 5111. **


# Question 9 

## part a

```{r,echo=FALSE}
m9 <- gam(death ~ s(pm10median) + s(o3median) + s(so2median) + s(tmpd),
          family = "poisson",
          data = new.chicago)

par(mfrow = c(2,2), mar = c(4,4,2,1))
plot(m9,
     residuals = TRUE,
     pch = 16,
     cex = 0.35, 
     main = "with residuals")

par(mfrow = c(2,2), mar = c(4,4,2,1))
plot(m9, residuals = FALSE, lwd = 2, main = "without residuals")

par(mfrow = c(2,2), mar = c(4,4,2,1))


```
**The partial response function for pm10median shows a generally increasing trend with the smooth rising gradually at higher particle levels indicating higher expected deaths as PM10 increases. The o3median smooth is relatively flat across most values but increases exponentially at higher ozone concentrations. The so2median smooth shows a weak nonlinear pattern remaining flat for lower so2 levels and dips around a concentration of 10 and then rapidly increases afterwards. Lastly, the tmpd smooth shows a nonlinear shape decreasing over moderate temperatures and then rising sharply at high temperatures at around 75 degrees farheinheit.**

## part b

```{r,echo=FALSE}

plot(new.chicago$date, new.chicago$death,
     pch = 16, cex = 0.5, col = "grey40",
     xlab = "date", ylab = "death",
     main = "fitted values as a function of time")


lines.mdl.NApadded(m9, data = new.chicago, xcol = "date",
                   col = "red", lwd = 2)

legend("topright",
       legend = c("Observed deaths", "GAM fitted mean"),
       col = c("grey40", "red"),
       pch = c(16, NA),
       lwd = c(NA, 2),
       bty = "n")
```
**The outliers are still present. The largest spikes in death is still around 1995 which remain far above the gam fitted mean. Even though the time -averaged predictors produces a slightly smoother fitted curve, it does not eliminate extreme outliers.**


# Question 10

## part a

```{r,echo=FALSE,message=FALSE}
library(dplyr)
library(knitr)
new.chicago$date <- as.Date(new.chicago$date)
out_idx  <- which.max(new.chicago$death)
out_date <- new.chicago$date[out_idx]
win_dates <- seq(out_date - 3, out_date + 3, by = "day")
prev_date <- as.Date(format(out_date, "%Y-%m-%d"))
prev_year_dates <- seq(prev_date - 365 - 3,
                       prev_date - 365 + 3,
                       by = "day")
window_now <- new.chicago %>%
  filter(date %in% win_dates) %>%
  arrange(date) %>%
  mutate(period = "Outlier Window")

window_prev <- new.chicago %>%
  filter(date %in% prev_year_dates) %>%
  arrange(date) %>%
  mutate(period = "Same Window (1 Year Earlier)")
comparison_table <- bind_rows(window_now, window_prev)

kable(comparison_table, digits = 3)
```

**Looking at the rows corresponding to the largest death counts, temperature is extremely high with values around 85-89 degrees Fahrenheit which are near the upper extreme of the dataset. Ozone is also high around 21-29 compared to its mean of -0.0078. In contrast PM10 and SO2 does not show an unusual spike in these rows. **

## part b

```{r,echo=FALSE}
m11 <- gam(death ~ 
             s(pm10median) +
             s(so2median) +
             te(tmpd, o3median),   # interaction
           family = "poisson",
           data = new.chicago)
par(mfrow = c(2,3), mar = c(4,4,2,1))

plot(m11,
     residuals = TRUE,
     pch = 16,
     cex = 0.35,
     lwd = 2,
     main = "with residuals")


par(mfrow = c(2,3), mar = c(4,4,2,1))

plot(m11,
     residuals = FALSE,
     lwd = 2,
     main = "without residuals")

```
**Fitting a model with an interaction between temperature and ozone is sensible since ozone formation is strongly temperature-dependent as the heat waves may amplify pollution-related deaths. The previous problem shows that mortality risk increases when both temperature and ozone are high indicating a synergy rather than being purely additive**

## part c

```{r,echo=FALSE}
m12 <- gam(death ~ 
             s(pm10median) +
             s(so2median) +
             te(tmpd, o3median),
           family = "poisson",
           data = new.chicago,
           na.action = na.exclude)
plot(new.chicago$date, new.chicago$death,
     pch = 16, cex = 0.5, col = "grey40",
     xlab = "date", ylab = "death",
     main = "fitted values vs time")

lines(new.chicago$date, fitted(m12),
      col = "red", lwd = 2)

legend("topright",
       legend = c("Observed deaths", "GAM fitted mean"),
       col = c("grey40", "red"),
       pch = c(16, NA),
       lwd = c(NA, 2),
       bty = "n")
```
**After including the interaction term, the fitted values show greater increases during extreme outliers particularly around 1995. As a result the most prominent outliers are somewhat reduced relative to the fitted curve but aren't eliminated as the largest death spike is clearly visible above the GAM fitted mean showing that it does not fully explain the most extreme death events. **

# Question 11

## part a

$E[R] = E[\frac{Y-\mu}{\sqrt \mu}] = \frac{1}{\sqrt \mu} (E[Y]-\mu)$

**Since we know that for a poisson distribution $E[Y] = \mu$**

$E[R] = \frac{1}{\sqrt \mu} (E[Y]-\mu) = \frac{1}{\sqrt \mu} (\mu-\mu) = 0$

$Var[R] = Var[\frac{Y-\mu}{\sqrt \mu}] = \frac{1}{\mu} Var(Y)$

**Since we know that for a poisson distribution $var[Y] = \mu$**

$Var[R] = \frac{1}{\mu} \mu = 1$


## part b

```{r,echo=FALSE}
rP <- residuals(m12, type = "pearson")
mu <- fitted(m12)

plot(mu, rP, pch=16, cex=0.4, col="grey40",
     xlab="Fitted values", ylab="Pearson residuals",
     main = "fitted values vs pearson residuals")
abline(h=0, lwd=2)

par(mfrow=c(2,2))
for (v in c("pm10median","o3median","so2median","tmpd")) {
  plot(new.chicago[[v]], rP,
       pch=16, cex=0.4, col="grey40",
       xlab=v, ylab="Pearson residuals",
       main = "partial response vs person residuals")
  abline(h=0, lwd=2)
}

par(mfrow=c(1,1))
```
**The residual plots suggest that the pearson residuals center around zero as shown in our plots where the points fluctuate above and below the horizontal 0 line without systematic shift. Although there is some variation in certain regions, there is no clear evidence that the residuals consistently deviate from zero across the range of fitted values or predictors. This indicates that the residuals have mean approximately zero overall. However, we do not expect them to have mean exactly zero  since E[R]=0 applies to the distribution as a whole and not individually.**

## part c

```{r,echo=FALSE}
rP_sq <- rP^2
mu <- fitted(m12)
plot(mu, rP_sq,
     pch = 16, cex = 0.4, col = "grey40",
     xlab = "Fitted values",
     ylab = "Squared Pearson residuals",
     main = "fitted values vs squared perason residuals")

abline(h = 1, lwd = 2)   
par(mfrow = c(2,2))

for (v in c("pm10median","o3median","so2median","tmpd")) {
  plot(new.chicago[[v]], rP_sq,
       pch = 16, cex = 0.4, col = "grey40",
       xlab = v,
       ylab = "Squared Pearson residuals",
       main = "partial response vs squared p.residuals")
  abline(h = 1, lwd = 2)
}

par(mfrow = c(1,1))
```
**The squared pearson plot does not show mean exactly equal to 1 everywhere. While many values lie near 1, there is substantial variability and several large spikes and extreme fitted values. This can be seen by the wide vertical spread and presence of large residual values rather than points closely clustered around 1. The Squared pearson residuals are expected to have approximately 1 overall under a correct poisson model, however, not at every value of a predictor and thus we expect the residuals to be centered around 1 overall, however, are not expected to equal 1 uniformely across all predictor values.**

## part d

**The results of Question 11(a) are theoretical benchmarks for the residual diagnostics in Questions 11(b) and 11(c). From 11 a we know that under a correctly specified poisson model, pearson residuals are expected to have mean zero and variance one. Along with this, the residual plots in 11(b) should show values centered around zero, while the squared Pearson residuals in 11(c) fluctuates around one. Moreover, severe deviation from mean zero or mean one may indicate model misspecification or violations of the Poisson variance assumption, such as over dispersion.**

# Question 12

## part a

**around 4 hours**

## part b

**i,ii, and v are true**

## part c

**ii**

## part d

**i.ii.and iv**
