
---
title: "36-402 Homework 4"
subtitle: Due Tuesday, Feburary 12th by 5:59pm
author: "Platt Yoon"
output: pdf_document
urlcolor: blue
---


# Problem 1:

```{r,echo=FALSE}
library(knitr)
df <- read.csv("nampd.csv")
d <- subset(df, !is.na(ln_old_mass) & !is.na(ln_mass))
fit <- lm(ln_mass ~ ln_old_mass, data = d)
d$anc_g <- exp(d$ln_old_mass)
d$new_g <- exp(d$ln_mass)
coef_vec <- coef(fit)
a <- coef(fit)[1]
b <- coef(fit)[2]
xgrid <- seq(min(d$anc_g), max(d$anc_g), length.out = 300)
ygrid_lm <- exp(a) * xgrid^b

plot(d$anc_g, d$new_g,
     xlab = "Ancestral mass (grams)",
     ylab = "New mass (grams)",
     pch = 16, cex = 0.6)
coef_df <- data.frame(
  Parameter = c("Intercept", "Slope"),
  Estimate = as.numeric(coef(fit))
)
lines(xgrid, ygrid_lm, lwd = 2)       

legend("topleft",
       legend = c("linear regression on logs"),
       lwd = 2, bty = "n")

knitr::kable(coef_df, digits = 4)
```
**The slope tells us that a 1 percent increase in ancestral mass is associated with 0.995 percent increase in new mass on average. Since the slope is close to 1 it means that descendant mass is almost proportional to ancestral mass with very weak regression toward the mean on the multiplicative scale. The intercept is 0.1411 when we exponent this($e^{0.1411} = 1.15$) we can interpret this as when the ancestral mass is 1gram($e^0 =1$), the predicted new mass is about 1.15 grams before noise.**

# Problem 2: 

```{r,echo=FALSE}
fit_spline <- smooth.spline(d$ln_old_mass, d$ln_mass)
pred_spline <- predict(fit_spline, x = log(xgrid))
ygrid_spline <- exp(pred_spline$y)
plot(d$anc_g, d$new_g,
     xlab = "Ancestral mass (grams)",
     ylab = "New mass (grams)",
     pch = 16, cex = 0.6)

lines(xgrid, ygrid_lm, lwd = 2, col = "blue")       
lines(xgrid, ygrid_spline, lwd = 2, col = "red")   

legend("topleft",
       legend = c("linear on logs", "smoothing spline on logs"),
       lwd = 2, bty = "n")

```


# Problem 3:
```{r,echo=FALSE}
#Directly from chapter 6 notes(code)
resample.data.frame <- function(data) {
  sample.rows <- resample(1:nrow(data))
  return(data[sample.rows,])
}
equitails <- function(x, alpha) {
  lower <- quantile(x, alpha/2)
  upper <- quantile(x, 1-alpha/2)
  return(c(lower, upper))
}
bootstrap.se <- function(statistic, simulator, B) {
    bootstrap(rboot(statistic, simulator, B), summarizer=sd)
}
rboot <- function(statistic, simulator, B) {
  tboots <- replicate(B, statistic(simulator()))
  if(is.null(dim(tboots))) {
      tboots <- array(tboots, dim=c(1, B))
  }
  return(tboots)
}
resample <- function(x) { sample(x,size=length(x),replace=TRUE) }

bootstrap.ci <- function(statistic=NULL, simulator=NULL, tboots=NULL,
                         B=if(!is.null(tboots)) { ncol(tboots) },
                         t.hat, level) {
  if (is.null(tboots)) {
    stopifnot(!is.null(statistic))
    stopifnot(!is.null(simulator))
    stopifnot(!is.null(B))
    tboots <- rboot(statistic, simulator, B)
  }
  alpha <- 1-level
  intervals <- bootstrap(tboots, summarizer=equitails, alpha=alpha)
  upper <- t.hat + (t.hat - intervals[,1])
  lower <- t.hat + (t.hat - intervals[,2])
  CIs <- cbind(lower=lower, upper=upper)
  return(CIs)
}

bootstrap <- function(tboots, summarizer, ...) {
  summaries <- apply(tboots, 1, summarizer, ...)
  return(t(summaries))
}


B <- 800   
xgrid_ln <- log(xgrid)
main.curve <- predict(fit_spline, xgrid_ln)$y
resample.residuals.mass <- function() {
  new.frame <- d
  fitted_vals <- predict(fit_spline, d$ln_old_mass)$y
  new.frame$ln_mass <- fitted_vals + resample(d$ln_mass - fitted_vals)
  return(new.frame)
}

resample.cases.mass <- function() { resample.data.frame(d) }
spline.curve <- function(data) {
  mdl <- smooth.spline(x=data$ln_old_mass, y=data$ln_mass)
  return(predict(mdl, xgrid_ln)$y)
}

#(a)
boot.curves.resid <- rboot(statistic=spline.curve,
simulator=resample.residuals.mass,B=B)

ci.resid.95 <- bootstrap.ci(tboots=boot.curves.resid,
t.hat=main.curve, level=0.95)
#(b) 
se.resid <- bootstrap.se(statistic=spline.curve,
simulator=resample.residuals.mass, B=B)
se.resid <- as.numeric(se.resid) 
lower.resid.se <- main.curve - 2*se.resid
upper.resid.se <- main.curve + 2*se.resid

#(c) 
boot.curves.case <- rboot(statistic=spline.curve,
simulator=resample.cases.mass, B=B)

ci.case.95 <- bootstrap.ci(tboots=boot.curves.case,
t.hat=main.curve,level=0.95)

#(d)
se.case <- bootstrap.se(statistic=spline.curve,
simulator=resample.cases.mass,B=B)

se.case <- as.numeric(se.case)
lower.case.se <- main.curve - 2*se.case
upper.case.se <- main.curve + 2*se.case

plot(exp(d$ln_old_mass), exp(d$ln_mass),
     main = "Question 3 (a)-(d)",
     xlab="Ancestral mass (grams)",
     ylab="New mass (grams)",
     pch=16, cex=0.6)
#Q2
lines(xgrid, exp(main.curve), lwd=2)

#(a)
lines(xgrid, exp(ci.resid.95[,"lower"]), lwd=2, col="blue")
lines(xgrid, exp(ci.resid.95[,"upper"]), lwd=2, col="blue")

#(b)
lines(xgrid, exp(lower.resid.se), lwd=2, col="red")
lines(xgrid, exp(upper.resid.se), lwd=2, col="red")

#(c)
lines(xgrid, exp(ci.case.95[,"lower"]), lwd=2,col="green")
lines(xgrid, exp(ci.case.95[,"upper"]), lwd=2,col="green")

#(d)
lines(xgrid, exp(lower.case.se), lwd=2,col="yellow")
lines(xgrid, exp(upper.case.se), lwd=2,col="yellow")

legend("topleft",legend = c("Spline regression",
                            "Residual 95% CI",
                            "Residual ±2 SE",
                            "Case 95% CI",
                            "Case ±2 SE"),
       lwd = 2,
       col = c("black","blue", "red", "green", "yellow"),
       bty = "n")


```


## PART A


## PART B

## PART C

## PART D

## PART E
**For the resampling of residuals, the $95%$ confidence bands are almost identical to the ±2 standard error bands, in which there are slight deviation at more spare-data regions. On the other hand, the resampling cases method produced different $95%$ confidence bands and ±2 standard error bands.In both cases neither should necessarily match since the ±2 standard error bands are a normal-approximation and assumes that each grid point is roughly normally distributed, while the $95%$ confidence band is a bootstrap interval where bootstrap distribution for a spline can be skewed and non-symmetric.**

# Problem 4:

```{r}
xmin<-1.8
xmax<-1e15
sigma2<-0.63
rmass <- function(XA, r, sigma2=0.63, xmin=xmin, xmax=xmax) {
  sigma <- sqrt(sigma2)
  XD <- 0
  while ((XD < xmin) || (XD > xmax)) {
    mu <- predict(r, x = log(XA))$y     
    Z <- rnorm(1, mean=0, sd=sigma)
    XD <- exp(mu + Z)                  
  }
  return(XD)
}

```


## Part A

**rmass would need to have a total of 5 variables: Ancestral mass(in grams) X_A, Minimum allowed species mass(grams) xmin, maximum allowed species mass(grams) xmax, Regression function (estimated from question 2), and Variance($\sigma^2$)**

## Part B

```{r,echo=FALSE,results= 'hide'}
XA_low  <- xmin * 1.01
XA_high <- xmax / 1.01
XD_low  <- replicate(B, rmass(XA=XA_low,  r=fit_spline, sigma2=sigma2, xmin=xmin, xmax=xmax))
XD_high <- replicate(B, rmass(XA=XA_high, r=fit_spline, sigma2=sigma2, xmin=xmin, xmax=xmax))
all((XD_low >= xmin) & (XD_low <= xmax))
all((XD_high >= xmin) & (XD_high <= xmax))
```
**The output is always between xmin and xmax even when $X_A$ is brought near either limit when checking them through random sampling of data points.**

## PART C


```{r,echo=FALSE}
set.seed(1)

XA <- exp(seq(log(xmin), log(xmax), length.out = 150))
XD <- sapply(XA, function(x) rmass(XA=x, r=fit_spline, sigma2=sigma2, xmin=xmin, xmax=xmax))
fit_spline_sim <- smooth.spline(x = log(XA), y = log(XD))

orig_curve_ln <- main.curve
sim_curve_ln  <- predict(fit_spline_sim, xgrid_ln)$y

plot(exp(d$ln_old_mass), exp(d$ln_mass),
     xlab="Ancestral mass (grams)",
     ylab="New mass (grams)",
     pch=16, cex=0.6)

lines(xgrid, exp(orig_curve_ln), lwd=2)
lines(xgrid, exp(sim_curve_ln),  lwd=2, lty=2)

legend("topleft",
       legend=c("Original regression curve", "Refit curve on simulated data"),
       lwd=2, lty=c(1,2), bty="n")

```
**The original regression curve and the refit curve on simulated data are not identical. they shouldn't be identical since each $X_D$ includes new random noise, and rmass also rejects values outside of xmin and xmax, and therefore, the refit spline essentially is estimating the same underlying curve with additional noise with fairly few points(150)**

# Problem 5:
```{r}
origin <- function(masses, r, sigma2=0.63, xmin=1.8, xmax=1e15) {
  i <- sample(1:length(masses), size=1)      
  XA <- masses[i]
  d1 <- rmass(XA=XA, r=r, sigma2=sigma2, xmin=xmin, xmax=xmax)
  d2 <- rmass(XA=XA, r=r, sigma2=sigma2, xmin=xmin, xmax=xmax)
  masses[i] <- d1
  return(c(masses, d2))
}
```


## Part A
```{r,echo=FALSE}
set.seed(1)

B <- 5000
XA <- 120
out <- replicate(B, origin(masses=c(XA), r=fit_spline), simplify="matrix")
d1 <- out[1, ]
d2 <- out[2, ]

matches_ancestor <- c(any(d1 == XA), any(d2 == XA))

knitr::kable(
  data.frame(
    Any_Exact_Match_To_XA = matches_ancestor,
    row.names = c("Descendant 1", "Descendant 2")
  )
)

ks_marginals <- ks.test(d1, d2)

knitr::kable(
  data.frame(
    Test = "KS test: Descendant 1 vs Descendant 2",
    p_value = as.numeric(ks_marginals$p.value)
  ),
  check.names = FALSE
)

corr_val <- cor(d1, d2)

knitr::kable(
  data.frame(`Correlation between two descendants` = corr_val),
  check.names = FALSE
)


```
**Neither descendant mass should equal the ancestral mass since the noise term is drawn from a continuous normal distribution and thus the probability that the noise term being exactly 0 is close to impossible. Based on the results, neither component matches the ancestral mass and both components have the roughly the same marginal distribution since our two-sample KS test comparing the two descendant gives p = 0.4354. This means that there is no evidence that descendant 1 and descendant 2 come from different distributions which is what we expect since the model generates two descendants symmetrically so their marginal distributions should match. Lastly, since the estimated correlation is -0.00977, and is extremely close to 0, it is consistent that the two descendants are approximately noncorrelated.**


## Part B

```{r,echo=FALSE}
set.seed(1)
B <- 2000
m_vals <- c(5, 20) 

len_check <- sapply(m_vals, function(m) {
  masses <- rep(120, m)  
  out_lens <- replicate(B, length(origin(masses=masses, r=fit_spline)))
  all(out_lens == m + 1)
})

knitr::kable(
  data.frame(m = m_vals, `Is the result 1 greater than m` = as.logical(len_check))
)

```


## Part C
```{r,echo=FALSE}
set.seed(1)

check_match <- function(m) {
  masses <- seq(10, by=10, length.out=m)  

  results <- replicate(B, {
    out <- origin(masses = masses, r = fit_spline)
    sum(is.element(masses, out))
  })
  all(results == (m - 1))
}
match_check <- sapply(m_vals, check_match)
tmp <- data.frame(m = m_vals, as.logical(match_check))
colnames(tmp) <- c("m", "Do m-1 entries match input?")
knitr::kable(tmp)

```



# Problem 6

```{r}
extinct.prob <- function(masses, rho, beta) {
  return(beta * masses^rho)
}
```


## Part A

```{r,echo=FALSE}
masses <- c(100, 1600, 10000)
rho <- 1/2
beta <- 1/200

out <- extinct.prob(masses, rho=rho, beta=beta)
expected <- beta * sqrt(masses)

knitr::kable(
  data.frame(
    mass = masses,
    output = out,
    expected = expected,
    difference = out - expected
  )
)

```



## Part B
```{r,echo=FALSE}
set.seed(1)
masses <- c(1, 10, 100, 1000, 1e6)
beta <- 1/200
rho <- 0
out <- extinct.prob(masses, rho=rho, beta=beta)
knitr::kable(
  data.frame(
    mass = masses,
    prob = out,
    equals_beta = (out == beta)
  )
)
```



## PART C

```{r,echo=FALSE}
tests <- list(
  list(mass=100,   rho=0.5,   beta=1/200),
  list(mass=5000,  rho=0.025, beta=1/100),
  list(mass=2e6,   rho=2,     beta=1/300)
)

equal_output_check <- sapply(tests, function(tst) {
  masses <- rep(tst$mass, 5)
  probs <- extinct.prob(masses, rho=tst$rho, beta=tst$beta)
  length(unique(probs)) == 1
})

knitr::kable(
  data.frame(
    test = 1:length(tests),
    mass = sapply(tests, function(tst) tst$mass),
    rho  = sapply(tests, function(tst) tst$rho),
    beta = sapply(tests, function(tst) tst$beta),
    outputs_all_equal = as.logical(equal_output_check)
  )
)

```


## PART D

```{r,echo=FALSE}
masses <- c(100, 1600, 10000, 25000)
rho <- 0.5
beta <- 1/200

probs <- extinct.prob(masses, rho=rho, beta=beta)

all_distinct <- (length(unique(probs)) == length(probs))
knitr::kable(
  data.frame(`All probabilities distinct?` = all_distinct),
  check.names = FALSE
)

```


# Problem 7

```{r}
extinction <- function(masses, rho, beta) {
  n <- length(masses)
  if (n == 0) return(masses)
  p <- extinct.prob(masses, rho=rho, beta=beta)
  extinct_flag <- rbinom(n, size=1, prob=p)
  survivors <- masses[extinct_flag == 0]
  return(survivors)
}
```


## Part A
```{r,echo=FALSE}
set.seed(1)
masses <- c(100, 500, 2000, 1e5)
rho <- 0.5
beta <- 0
B <- 2000
same_as_input <- replicate(B, identical(extinction(masses, rho=rho, beta=beta), masses))
knitr::kable(
  data.frame(`Output always equal to input?` = all(same_as_input))
)

```



## Part B
```{r,echo=FALSE}
set.seed(1)
m <- 1000
x0 <- 100
masses <- rep(x0, m)
rho <- 0.5
beta <- 0.5 / (x0^rho)   
B <- 2000
out_lengths <- replicate(B, length(extinction(masses, rho=rho, beta=beta)))

knitr::kable(
  data.frame(
    input_length = m,
    mean_output_length = mean(out_lengths),
    expected_half = m/2
  )
)

```



## Part C

```{r,echo=FALSE}
set.seed(1)
B <- 2000
all_from_input <- replicate(B, {
  out <- extinction(masses, rho=rho, beta=beta)
  all(is.element(out, masses))
})

knitr::kable(
  data.frame(`All survivors were in the input?` = all(all_from_input))
)
```



# Problem 8:
```{r}
evolve_step <- function(masses, r, sigma2=0.63, xmin=1.8, xmax=1e15,
                        rho=0.025, beta=1/5000) {
  masses <- origin(masses, r=r, sigma2=sigma2, xmin=xmin, xmax=xmax)
  masses <- extinction(masses, rho=rho, beta=beta)

  return(masses)
}
```
**The evolve_step() works since it is the composition of the function origin and extinction. From the previous problem, we already identified that origin performs correctly and increases the vector length by 1 while keeping masses within bounds. From questions 6 and 7 we know that the extinction probabilities are computed correctly and extinction removes species according to those probabilities without introducing new values.**

# Problem 9:
```{r}
mass_evolve <- function(masses, r, T, sigma2=0.63, xmin=1.8, xmax=1e15,
                        rho=0.025, beta=1/5000) {
  for (t in 1:T) {
    masses <- evolve_step(masses, r=r, sigma2=sigma2, xmin=xmin, xmax=xmax,
                          rho=rho, beta=beta)
  }
  return(masses)
}
```
**mass_evolve is a direct iterative extension of evolve_step where the function iterates exactly T  times and applies evolve_step repeatedly. From problem 8 we already verified that evolve_step correctly performs one full model update while keeping invariant such as valid bounded mass. When T=0, the loop executes zero times and returns the original input vector unchanged and when T =1,, the output is identical to a single call in evolve_step() which shows that the loop structure behaves correctly. Therefore, since each iteration preserves mass bounds and model structure we can safely say that the function correctly runs.**

# Problem 10:

## PART A
```{r,echo=FALSE}
set.seed(1)
masses.1 <- mass_evolve(masses=c(120), r=fit_spline, T=2e5)

hist(log(masses.1), xlab="Mass (grams)", main="log of masses.1")

```


## PART B
```{r,echo=FALSE}
set.seed(2)
masses.2 <- mass_evolve(masses=c(120), r=fit_spline, T=2e5)
hist(log(masses.2), xlab="Mass (grams)", main="log(masses.2)")
```

## PART C
```{r,echo=FALSE}
set.seed(3)
masses.3 <- mass_evolve(masses=c(120), r=fit_spline, T=4e5)
hist(log(masses.3), xlab="Mass (grams)", main="log(masses.3)")
```

## PART D
```{r,echo=FALSE}
set.seed(4)
masses.4 <- mass_evolve(masses=c(40, 1000), r=fit_spline, T=2e5)
hist(log(masses.4), xlab="Mass (grams)", main="log(masses.4)")

```


```{r}
set.seed(5)
masses.5 <- mass_evolve(masses=c(40, 1000), r=fit_spline, T=2e5)
hist(log(masses.5), xlab="Mass (grams)", main="log(masses.5)")

```

## PART E
```{r,echo=FALSE}
par(mfrow=c(2,2))
qqplot(log(masses.1), log(masses.2),
       xlab="log(masses.1)", ylab="log(masses.2)",
       main="masses.1 vs masses.2"); abline(0,1)

qqplot(log(masses.1), log(masses.3),
       xlab="log(masses.1)", ylab="log(masses.3)",
       main="masses.1 vs masses.3"); abline(0,1)

qqplot(log(masses.4), log(masses.5),
       xlab="log(masses.4)", ylab="log(masses.5)",
       main="masses.4 vs masses.5"); abline(0,1)

qqplot(log(masses.1), log(masses.4),
       xlab="log(masses.1)", ylab="log(masses.4)",
       main="masses.1 vs masses.4"); abline(0,1)

par(mfrow=c(1,1))
```
**All Q–Q plots are fairly linear, indicating similar distributions across simulation runs. Most importantly, deviations are primarily visible in the upper tails and the q-q plots masses.1 vs masses.4 and masses.1 vs masses.3  shows the most noticeable deviations, reflecting the effect of different initial conditions as well as increasing the simulation time alters tail behaviour.**

# Problem 11:

## PART A

```{r,echo=FALSE}
library(dplyr)
mom <- read.table("https://www.stat.cmu.edu/~cshalizi/uADA/26/hw/04/MoM.txt",
                  header = TRUE)
mom <- mom %>% filter(mom$land == 1)
hist(log(mom$mass),
     xlab = "log Mass (grams)",
     main = "Histogram of log(Mammal Masses)")
```
**The histogram of log(mammal masses) for land species show a strongly right skewed distribution with many small-bodied mammals and relatively few large bodied species. This shape closely resembles the distributions produced by the simulator. The main differences are that the simulated distributions whow heavier concentrations for lower log Mass than that of the MoM histogram. Moreover, the simulated histograms show higher skewness than that of the MoM histogram.**

## PART B

## PART C
```{r,echo=FALSE}
names(mom)
mom.mass <- mom$mass
mom.land.mass <- mom.mass[mom$land == 1]
qqplot(log(mom.land.mass), log(masses.1),
       xlab="log(land masses from MoM)",
       ylab="log(simulated masses.1)",
       main="Land (MoM) vs masses.1")
abline(0,1)
qqplot(log(mom.land.mass), log(masses.2),
       xlab="log(land masses from MoM)",
       ylab="log(simulated masses.2)",
       main="Land (MoM) vs masses.2")
abline(0,1)
qqplot(log(mom.land.mass), log(masses.3),
       xlab="log(land masses from MoM)",
       ylab="log(simulated masses.3)",
       main="Land (MoM) vs masses.3")
abline(0,1)
qqplot(log(mom.land.mass), log(masses.4),
       xlab="log(land masses from MoM)",
       ylab="log(simulated masses.4)",
       main="Land (MoM) vs masses.4")
abline(0,1)
qqplot(log(mom.land.mass), log(masses.5),
       xlab="log(land masses from MoM)",
       ylab="log(simulated masses.5)",
       main="Land (MoM) vs masses.5")
abline(0,1)

```



# Problem 12:

**Overall, the simulation model fairly matches the broad qualitative shapes of land-mammal mass distribution where on the log scale both are strongly right skewed and the Q-Q plots comparing MoM land mass to each simulated runs are all close to linear especially throughout the middle quantities. However, the model does not match reality perfectly and the differences between the model and reality are larger and more systematic. Across the simulation vs simulation Q-Q plots in 10e the runs are very close to each other and mostly differ only by small tail noises, whereas the MoM vs simulation QQ plots show a consistent pattern that the upper tail points bend away from the 45 degree line meaning that the simulation tend to produce more extreme large masses than the real land-mammal data. In terms of qualitative distinctions between simulation to simulation and simulation to reality, the simulation to simulation differences tends to be mostly random variation and small deviations concentrated in the tails with no strong systematic patterns while the simulation to reality differences show systematic tail deviations and simulations often show heavier upper tails reflecting model limitations and not randomness. **

# Problem 13:

**It took me approximately 14 hours to complete**




# Extra Credit

```{r}
fitted_vals <- predict(fit_spline, d$ln_old_mass)$y
resid_spline <- d$ln_mass - fitted_vals
rmass_resid <- function(xA, r, resid, xmin=1.8, xmax=1e15) {
  mu <- predict(r, log(xA))$y
  z <- resample(resid)[1]       
  xD <- exp(mu + z)
  while ((xD < xmin) | (xD > xmax)) {
    z <- resample(resid)[1]
    xD <- exp(mu + z)
  }
  return(xD)
}

origin_resid <- function(masses, r, resid, xmin=1.8, xmax=1e15) {
  i <- sample(1:length(masses), size=1)
  xA <- masses[i]
  xD1 <- rmass_resid(xA, r=r, resid=resid, xmin=xmin, xmax=xmax)
  xD2 <- rmass_resid(xA, r=r, resid=resid, xmin=xmin, xmax=xmax)
  return(c(masses[-i], xD1, xD2))
}

set.seed(1)
masses_gauss <- mass_evolve(masses=c(120), r=fit_spline, T=2e5)

evolve_step_resid <- function(masses, r, resid, sigma2=0.63, xmin=1.8, xmax=1e15,
                              rho=0.025, beta=1/5000) {
  masses <- origin_resid(masses, r=r, resid=resid, xmin=xmin, xmax=xmax)
  masses <- extinction(masses, rho=rho, beta=beta)
  return(masses)
}

mass_evolve_resid <- function(masses, r, resid, T, xmin=1.8, xmax=1e15,
                              rho=0.025, beta=1/5000) {
  for (t in 1:T) {
    masses <- evolve_step_resid(masses, r=r, resid=resid,
                                xmin=xmin, xmax=xmax, rho=rho, beta=beta)
  }
  return(masses)
}

set.seed(1)
masses_resid <- mass_evolve_resid(masses=c(120), r=fit_spline, resid=resid_spline, T=2e5)
mom_land <- mom 

qqplot(log(mom_land$mass), log(masses_gauss),
       xlab="log(MoM land masses)", ylab="log(simulated masses)",
       main="MoM vs Gaussian-noise simulation"); abline(0,1)

qqplot(log(mom_land$mass), log(masses_resid),
       xlab="log(MoM land masses)", ylab="log(simulated masses)",
       main="MoM vs residual-resampling simulation"); abline(0,1)
```

**In our current model the descendent mass is generated on a log scale such that $logX_D = rlog(X_A) + Z,$ where Z is sampled such that $Z~N(0,\sigma^2)$ and therefore the only thing to do is change how we generate Z by using resampling the residuals of the estimated regression curve by using resample(residuals_from_fit) from $logX-\hat{r}(logX)$.This requires computing residuals from the fitted regression curve and changing the descendant generator to use resample rather than rnorm and change corresponding functions accordingly**

**In terms of effects, the overall distributional shape remains similar under both methods based on the QQ plot and the main difference appears in the upper tails. Residual sampling produces a curve that tracks the MoM quantiles more closely at large masses while Gaussian noise generates more extreme high end values that are visible with stronger upward deviation in the upper tail.**

**Based on the Q-Q plot, the residual resampling QQ plot lies closer to the 45 degree line and smaller deviation in the upper tail compared to that of the gaussian noise QQ plot and therefore, indicates a better fit than the gaussian noise version.**



